Supplementary Material for ``Provisional Title''

### 2021-04-16 draft/supp.tex
more supp

% packages
\usepackage[utf8]{inputenc} % allows usage of spanish special characters
\usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
\usepackage{amsmath} % math expressions
\usepackage{upgreek} % upright greek letters for math
\usepackage{txfonts} % nice fonts 
\usepackage{authblk} % author affiliations
\usepackage{graphicx} % images
\usepackage{float} % image positioning
\usepackage{multirow} % allows merging cells in tables
\usepackage{mathtools} % nice matrices
\usepackage{rotating} % rotated text
\usepackage{caption} % customization of figure/table captions
\usepackage{subcaption} % collages of multiple images
\usepackage{hyperref} % hyperlinks
\usepackage{nameref} % cross-references between draft and supplementary material
\usepackage{zref-xr,zref-user} % more cross-references formatting
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage{xcolor} % text color
\usepackage[margin=3cm]{geometry} % margins
\usepackage[nottoc,numbib]{tocbibind} % add references to table of contents

@@

% customization
\setcounter{Maxaffil}{0} % affiliations
\renewcommand\Affilfont{\itshape\small} % style of affiliations text
\makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in reference list
\addto{\captionsenglish}{\renewcommand{\bibname}{Supplementary References}}
\newcommand{\idea}[1]{\textcolor{red}{#1}}
\newcommand{\mr}{\textcolor{red}{\textbf{[missing ref(s)]}}}
\newcommand{\Burl}[1]{\url{#1}}
\zxrsetup{toltxlabel} % cross references: draft & suppl mat
\zexternaldocument*[M-]{draft}
\hypersetup{
  colorlinks = true,
  citecolor=  black,
  linkcolor = {blue},
  filecolor = cyan % controls color of external ref, if used
}
\renewcommand{\thefigure}{\textbf{S\arabic{figure}}} % preceed the Suppl Mat figures with S
\renewcommand{\thetable}{\textbf{S\arabic{table}}} % preceed the Suppl Mat tables with S
\renewcommand{\theequation}{S\arabic{equation}} % preceed the Suppl Mat equations with S
\captionsetup[figure]{font=small, labelfont={bf}, labelformat={default},
   labelsep=period, name={Fig.}} % custom figure captions
\captionsetup[table]{font=small, labelfont={bf}, labelformat={default},
   labelsep=period, name={Table}} % custom table captions
\hypersetup{citecolor={blue}}

@@

\begin{equation}
d_T = \mathrm{Q}4 \left( \left\{d\right\} \right) + 
\mathrm{IQR} \left( \left\{d\right\} \right)
\label{supp-eqn:threshold}
\end{equation}
%
Where Q4 and IQR represent the fourth quantile and the interquartile range
respectively. Note that this threshold is analogous to the one used in
standard boxplots to identify outliers.

Finally, we discard the samples that are at a distance larger than $d_T$ from
all other replicates of the same community. These are replicate 2 of community
1 in glutamine and replicate 2 of community 8 in citrate (see Figure S1 \mr).
The remaining replicates were used in further analyses. We took the average
ESV abundance across replicates when indicated.

\subsubsection{Special cases}\label{supp-methods:data-proc:special}

\subsection{Inferring species abundance from sequencing data}
\label{supp-methods:species-from-seq}

@@

By sequencing each species individually, we can build a $M \times N$ 
matrix $\mathbf{Q}$ such
that the element in the $(i,j)$ position, $q_{ij}$, represents the frequency of
the $i$-th ESV when the $j$-th species is sequenced. $\mathbf{Q}$ can be used
to determine the fraction of sequences of a given ESV that are obtained from
sequencing a community with any arbitrary composition. In other words,
$\mathbf{Q}$ maps $\mathbf{x}$ to $\mathbf{s}$:

>>>>

% packages
\usepackage[utf8]{inputenc} % allows usage of spanish special characters
\usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
\usepackage{amsmath} % math expressions
\usepackage{upgreek} % upright greek letters for math
\usepackage{txfonts} % nice fonts 
\usepackage{authblk} % author affiliations
\usepackage{graphicx} % images
\usepackage{float} % image positioning
\usepackage{multirow} % allows merging cells in tables
\usepackage{makecell} % more table customization
\usepackage{mathtools} % nice matrices
\usepackage{rotating} % rotated text
\usepackage{caption} % customization of figure/table captions
\usepackage{subcaption} % collages of multiple images
\usepackage{hyperref} % hyperlinks
\usepackage{nameref} % cross-references between draft and supplementary material
\usepackage{zref-xr,zref-user} % more cross-references formatting
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage[table,xcdraw]{xcolor} % text and table background colors
\usepackage[margin=3cm]{geometry} % margins
\usepackage[nottoc,numbib]{tocbibind} % add references to table of contents

@@

% customization
\setcounter{Maxaffil}{0} % affiliations
\renewcommand\Affilfont{\itshape\small} % style of affiliations text
\makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in reference list
\addto{\captionsenglish}{\renewcommand{\bibname}{Supplementary References}}
\newcommand{\idea}[1]{\textcolor{red}{#1}}
\newcommand{\mr}{\textcolor{red}{\textbf{[missing ref(s)]}}}
\newcommand{\Burl}[1]{\url{#1}}
\zxrsetup{toltxlabel} % cross references: draft & suppl mat
\zexternaldocument*[M-]{draft}
\hypersetup{
  colorlinks = true,
  citecolor=  black,
  linkcolor = {blue},
  filecolor = cyan % controls color of external ref, if used
}
\renewcommand{\thefigure}{\textbf{S\arabic{figure}}} % preceed the Suppl Mat figures with S
\renewcommand{\thetable}{\textbf{S\arabic{table}}} % preceed the Suppl Mat tables with S
\renewcommand{\theequation}{S\arabic{equation}} % preceed the Suppl Mat equations with S
\captionsetup[figure]{font=small, labelfont={bf}, labelformat={default},
   labelsep=period, name={Fig.}} % custom figure captions
\captionsetup[table]{font=small, labelfont={bf}, labelformat={default},
   labelsep=period, name={Table}} % custom table captions
\hypersetup{citecolor={blue}}
\definecolor{lightgray}{gray}{0.9}

@@

\begin{equation}
d_T = \mathrm{Q}3 \left( \left\{d\right\} \right) + 
1.5 \; \mathrm{IQR} \left( \left\{d\right\} \right)
\label{supp-eqn:threshold}
\end{equation}
%
Where Q3 and IQR represent the third quartile and the interquartile range
respectively. Note that this threshold is analogous to the one used in
standard boxplots to identify outliers.

Finally, we discard the samples that are at a distance larger than $d_T$ from
all other replicates of the same community. These are replicate 2 of community
1 in glutamine and replicate 2 of community 8 in citrate (see Figure S1 \mr).
The remaining replicates were used in further analyses, taking the average
ESV abundance across replicates when indicated.

\subsubsection{Characterization of dominant species}
\label{supp-methods:data-proc:doms}

As described in the main text Methods \mr, generationally stabilized communities
were plated and the most abundant species
(determined by counting of colony morphotypes) was isolated
and allowed to regrow. We grew three biological
replicates of each isolated species in monoculture and then performed
16S rRNA gene sequencing. We characterize
the isolated species to, first,
determine whether multiple communities share the same dominant
and, second, to
make sure that the dominants were correctly identified.

Analogous to what we did in the section above
(\nameref{supp-methods:data-proc:community-composition}), we start by estimating
the sequencing error by comparing the distances between all pairs of replicates.
We obtain an ``error threshold'' like we did in equation \ref{supp-eqn:threshold},
and we discard replicates that fall at a larger distance
from all other replicates of the same isolate. This eliminates one replicate
from the monocultures of the isolates from glutamine communities 2 and 3, and
an additional replicate from citrate community 1.

We then compare each isolate with
one another. The distance between the vectors of ESV abundances of two
different isolates being below our threshold indicates that both
are the same species, or at the very least that we cannot
distinguish them through 16S sequencing. These instances were removed from 
the dominant pairwise competition analysis. We find that indeed some
communities share
the same (or undistinguishable) isolated dominant. This leaves us with 9 unique
isolates.

Table \ref{supp-tab:isolates} shows the relative ESV abundance of each unique
isolate, as well as which communities they were isolated from.
Table \mr shows the ESVs obtained from 16S sequencing
of the monocultures. Note that three of the isolates partially share a same
ESV; however, when growing them in pairwise competition there is always at least
another ESV that is unique to only one of them.
Further details can be found in the code for the analysis (see section 
\nameref{supp-datacode} of this Supplementary Material).

\begin{table}
\scriptsize
\centering
\rowcolors{2}{lightgray}{white}

\begin{tabular}{lccccccccc}
 & \makecell{Isolate 1 \\ G-1, C-8} 
 & \makecell{Isolate 2 \\ G-2, G-5, C-3, C-5 }
 & \makecell{Isolate 3 \\ G-3 }
 & \makecell{Isolate 4 \\ G-4, G-6 }
 & \makecell{Isolate 5 \\ G-7, G-8 }
 & \makecell{Isolate 6 \\ C-1 }
 & \makecell{Isolate 7 \\ C-2, C-7 }
 & \makecell{Isolate 8 \\ C-4 }
 & \makecell{Isolate 9 \\ C-6 } \\ \cline{2-10} 
\multicolumn{1}{l|}{Sequence 1} &  & 1 &  & 0.86 &  & 0.63 &  &  &  \\
\multicolumn{1}{l|}{Sequence 2} &  &  &  &  &  &  & 1 &  &  \\
\multicolumn{1}{l|}{Sequence 3} &  &  & 1 &  &  &  &  &  &  \\
\multicolumn{1}{l|}{Sequence 4} & 1 &  &  &  &  &  &  &  &  \\
\multicolumn{1}{l|}{Sequence 5} &  &  &  &  &  &  &  & 1 &  \\
\multicolumn{1}{l|}{Sequence 6} &  &  &  &  & 1 &  &  &  &  \\
\multicolumn{1}{l|}{Sequence 7} &  &  &  & 0.14 &  &  &  &  &  \\
\multicolumn{1}{l|}{Sequence 8} &  &  &  &  &  &  &  &  & 1 \\
\multicolumn{1}{l|}{Sequence 9} &  &  &  &  &  & 0.37 &  &  &
\end{tabular}

\caption{\textbf{Species to ESVs map of the dominant species isolated from
the communities.} For every isolate we indicate the community (or communities if
they share dominants) from which it was obtained. G-$n$ and C-$n$
($n=1 \; \text{to} \; 8$)
denote the $n$-th community in glutamine and citrate respectively.
Values in the table represent relative ESV abundances from 16S sequencing
of monocultures.
Empty entries of the table represent zeros.}
\label{supp-tab:isolates}

\end{table}

\begin{table}
\footnotesize
\centering
\rowcolors{0}{lightgray}{white}

\begin{tabular}{ll}

Sequence 1 &
\makecell[l]{
\tt GCAAGCGTTA ATCGGAATTA CTGGGCGTAA AGCGCGCGTA GGTGGTTTGT TAAGTTGGAT \\
\tt GTGAAAGCCC CGGGCTCAAC CTGGGAACTG CATCCAAAAC TGGCAAGCTA GAGTACGGTA \\
\tt GAGGGTGGTG GAATTTCCTG TGTAGCGGTG AAATGCGTAG ATATAGGAAG GAACACCAGT \\
\tt GGCGAAGGCG ACCACCTGGA CTGATACTGA CACTGAGGTG CGAAAGCGTG GGG
}

Sequence 2 &
\makecell[l]{
\tt GCAAGCGTTA ATCGGAATTA CTGGGCGTAA AGCGCACGCA GGCGGTCTGT CAAGTCGGAT \\
\tt GTGAAATCCC CGGGCTCAAC CTGGGAACTG CATTCGAAAC TGGCAGGCTA GAGTCTTGTA \\
\tt GAGGGGGGTA GAATTCCAGG TGTAGCGGTG AAATGCGTAG AGATCTGGAG GAATACCGGT \\
\tt GGCGAAGGCG GCCCCCTGGA CAAAGACTGA CGCTCAGGTG CGAAAGCGTG GGG
}

\end{tabular}

\caption{\textbf{Exact Sequence Variants of the isolated dominant species.}
Sequence numbering as in table \ref{supp-tab:isolates}.}
\label{supp-tab:esvs}

\end{table}

\subsection{Inferring species abundances from sequencing data}
\label{supp-methods:species-from-seq}

@@

By sequencing each species individually, we can build a $M \times N$ 
matrix $\mathbf{Q}$ such
that the element in the $(i,j)$ position, $q_{ij}$, represents the frequency of
the $i$-th ESV when the $j$-th species is sequenced (see table
\ref{supp-tab:isolates} and section \nameref{supp-methods:data-proc:doms} of
this Supplementary Material).
$\mathbf{Q}$ can be used
to determine the fraction of sequences of a given ESV that are generated when
sequencing a community with any arbitrary composition. In other words,
$\mathbf{Q}$ maps $\mathbf{x}$ to $\mathbf{s}$:
